Equalizer 2.0\\Convergence of a Parallel Rendering Framework

### 2017-04-03 paperIEEE.tex
Refine compound configs
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@22 db7c848b-7755-4df2-b763-8796797afa52
General review
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@23 db7c848b-7755-4df2-b763-8796797afa52

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Equalizer: A Mature Parallel Rendering Framework}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

@@

The continuing improvements in hardware integration lead to ever faster CPUs and GPUs, as well as higher resolution sensor and display devices. Moreover, increased hardware parallelism is applied in form of multi-core CPU workstations, massive parallel super computers, or cluster systems. Hand in hand goes the rapid growth in complexity of data sets from numerical simulations, high-resolution 3D scanning systems or bio-medical imaging, which causes interactive exploration and visualization of such large data sets to become a serious challenge. It is thus crucial for a visualization solution to take advantage of hardware accelerated scalable parallel rendering. In this systems paper we describe a new scalable parallel rendering framework called {\em Equalizer} that is aimed primarily at cluster-parallel rendering, but works as well in a shared-memory system. Cluster systems are the main focus  because workstation graphics hardware is developing faster than high-end graphics (super-) computers can absorb new developments, and also because clusters offer a better cost-performance balance.

Previous parallel rendering approaches typically failed in one of the following system requirements:
%
\begin{compactenum}\renewcommand{\labelenumi}{\alph{enumi})}
\item generic application support, instead of special domain solution
\item scalable abstraction of the graphics layer
\item exploit existing code infrastructure, such as proprietary scene graphs, molecular data structures, level-of-detail and geometry databases
\end{compactenum}

To date, generic and scalable parallel rendering frameworks  that can be adopted to a wide range of scientific visualization domains are not yet readily available. Furthermore, flexible
configurability to arbitrary cluster and display-wall configurations has also not been addressed in the past, but is of immense practical importance to scientists depending high-performance interactive visualization as a scientific tool. In this paper we present Equalizer, which is a novel flexible framework for parallel rendering that supports scalable performance, configuration flexibility, is {\em minimally invasive} with respect to adapting existing visualization applications, and is applicable to virtually any scientific visualization application domain.

The main contributions that Equalizer introduces in a single parallel rendering system, and which are presented in this paper are:
%
\begin{compactenum}\renewcommand{\labelenumi}{\roman{enumi})}
\item novel concept of compound trees for flexible configuration of graphics system resources,
\item easy specification of parallel task decomposition and image compositing choice through compound tree layouts,
\item automatic decomposition and distributed execution of rendering tasks according to compound tree,
\item support for parallel surface as well as transparent (volume) rendering through $z$-visibility as well as $\alpha$-blending compositing,
\item fully decentralized architecture providing network swap barrier (synchronization) and distributed objects functionality,
\item support for low-latency distributed frame synchronization and image compositing,
\item minimally invasive programming model.
\end{compactenum}

Equalizer is open source, available under the LGPL license from http://www.equalizergraphcis.com/, which allows it to be used both for open source and commercial applications. It is source-code portable, and has been tested on Linux, Microsoft Windows, and Mac OS X in 32 and 64 bit mode using both little endian and big endian processors.

%----------------------------------------------------------------------
\section{Related Work}\label{SEC_related}
%----------------------------------------------------------------------

@@

\begin{figure*}[ht]\center
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/dplex}
      \caption{\label{fig:dplex}Time-Multiplex}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/tile}
      \caption{\label{fig:tile}Tiles}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/pixel}
      \caption{\label{fig:pixel}Pixel}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/subpixel}
      \caption{\label{fig:subpixel}Subpixel}
  \end{subfigure}
  \caption{New Equalizer task decompositions for parallel rendering}
  \label{fig:compounds}
\end{figure*}

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to Equalizer
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in a multi-threaded,
multi-GPU configuration, the application needs to support running the rendering
threads asynchronously, as outlined in \sref{SEC_threading}. The output frame
rate of the destination channel is smoothened using a frame rate equalizer
(\sref{SEC_framerateEq}).

\begin{figure*}[ht]\center
  \begin{subfigure}[b]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  framerate_equalizer {}
  compound {
    channel "source1"
    phase 0 period 2
    outputframe  { name "frame" }
  }
  compound {
    channel "source2"
    phase 1 period 2
    outputframe  { name "frame" }
  }
  inputframe { name "frame" }
}
    \end{lstlisting}}
    \caption{\label{eqc:dplex}Time-Multiplex}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  outputtiles {
    name "queue"
    size [ 64 64 ]
  }
  compound {
    channel "destination"
    inputtiles { name "queue" }
  }
  compound {
    channel "source"
    inputtiles { name "queue" }
    outputframe {}
  }
  inputframe { name "frame.source" }
}
    \end{lstlisting}}
    \caption{\label{eqc:tiles}Tiles}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    pixel [ 0 0 2 1 ]
    outputframe { type texture }
  }
  compound {
    channel "source"
    pixel [ 1 0 2 1 ]
    outputframe{}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source" }
}
    \end{lstlisting}}
    \caption{\label{eqc:pixel}Pixel}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    subpixel [ 0 2 ]
    outputframe { type texture }
  }
  compound {
    channel "source"
    subpixel [ 1 2 ]
    outputframe{}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source" }
}
    \end{lstlisting}}
    \caption{\label{eqc:subpixel}Subpixel}
  \end{subfigure}
  \caption{Compound configurations for \fig{fig:compounds}}
  \label{eqc:compounds}
\end{figure*}

@@

\subsubsection{Frame Rate Equalizer}\label{SEC_framerateEq}

@@

\subsubsection{Thread Synchronization Modes}\label{SEC_threading}

@@

Head tracking is the minimal feature needed to support immersive
installations. Equalizer does support multiple, independent tracked views
through the observer abstraction introduced in \sref{SEC_observer}. Built-in
VRPN support enables the direct, application-transparent configuration of a VRPN
tracker device. Alternatively, an application can provide a $4\times 4$ tracking
matrix defining the transformation from the canvas coordinate system to the
observer. Both CAVE-like tracking with fixed projection surfaces and HMD
tracking modes are implemented.

@@

Applications can switch each view between mono and stereo rendering at runtime,
and can run both monoscopic and stereoscopic views concurrently
(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
does potentially involve the start and stop of resources and processes for
passive stereo or stereo-dependent task decompositions, as described in
\sref{SEC_reconfig}.

@@

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved
screens. These frusta describe a physically-correct display setup for a Virtual
Reality installation. \label{sec:swap} A canvas may have a software or hardware
swap barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a Collage barrier (\sref{SEC_barrier}) to synchronize all segments, the
swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
the swap command. The hardware swap barrier is implemented using the
\textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.

@@

\label{SEC_observer}
An observer looks at one or more views. It is described by the observer position
in the world and its eye separation. Each observer will have its own stereo
mode, focus distance and frame loop (framerate). This allows to have untracked
views and multiple tracked views, e.g., two HMDs, in the same application.

\subsection{Runtime Reconfiguration}\label{SEC_reconfig}

@@

\begin{compactdesc}
\item[Connection] A stream-oriented point-to-point communication
  line. The connections
  transmit raw data reliably between two endpoints for unicast connections, and
  between a set of endpoints for multicast connections. For unicast,
  process-local pipes, TCP and Infiniband RDMA are implemented. For multicast,
  a reliable, UDP-based protocol is discussed in \sref{SEC_RSP}.
\item[DataI/OStream] Abstracts the input and output of C++ data types from or to
  a set of connections by implementing output stream operators. Uses buffering
  to aggregate data for network transmission. Performs endian swapping during
  input if the endianness differs between the originating and local node.
\item[Node and LocalNode] The abstraction of a process in the cluster. Nodes
  communicate with each other using connections. A LocalNode listens on various
  connections and processes requests for a given process. Received data is
  wrapped in ICommands and dispatched to command handler methods. A Node is a
  proxy for a remote LocalNode. The Equalizer Client object is a LocalNode.
\item[Object] Provides object-oriented, versioned data distribution of C++
  objects between nodes. Objects are registered or mapped on a Local\-Node.
\end{compactdesc}

@@

\subsection{Reliable Stream Protocol}\label{SEC_RSP}

@@

The Collage Serializable implements one convenient usage pattern for object data
distribution. The \textsf{co::Serializable} data distribution is based on the
concept of dirty bits, allowing inheritance with data distribution. Dirty bits
form a 64-bit mask which marks the parts of the object to be distributed during
the next commit. For serialization, the application developer implements
\textsf{serialize} or \textsf{deserialize}, which are called with the bit mask
specifying which data has to be transmitted or received. During a commit or
sync, the current dirty bits are given, whereas during object mapping all dirty
bits are passed to the serialization methods.

@@

\subsection{Barriers, Queues and Object Maps}\label{SEC_barrier}

@@

Livre (Large-scale interactive volume rendering engine) is an out-of-core
parallel 4D volume renderer.

@@

@@

%----------------------------------------------------------------------
\section{Experimental Results}
%----------------------------------------------------------------------
\label{SEC_results}

@@

@@

%----------------------------------------------------------------------
\section{Discussion and Conclusion}
%----------------------------------------------------------------------
\label{SEC_conclusions}

@@

>>>>

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Equalizer 2.0\\Convergence of a Parallel Rendering Framework}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%
%
%\IEEEcompsocitemizethanks is a special \thanks that produces the bulleted
% lists the Computer Society journals use for "first footnote" author
% affiliations. Use \IEEEcompsocthanksitem which works much like \item
% for each affiliation group. When not in compsoc mode,
% \IEEEcompsocitemizethanks becomes like \thanks and
% \IEEEcompsocthanksitem becomes a line break with idention. This
% facilitates dual compilation, although admittedly the differences in the
% desired content of \author between the different types of papers makes a
% one-size-fits-all approach a daunting prospect. For instance, compsoc
% journal papers have the author affiliations above the "Manuscript
% received ..."  text while in non-compsoc journals this is reversed. Sigh.

@@

Highlight integration of many features into a general framework vs one-off
research implemenation of one algorithm.

%----------------------------------------------------------------------
\section{Related Work}\label{sec:related}
%----------------------------------------------------------------------

@@

\begin{figure*}[ht]\center
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/dplex}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/tile}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/pixel}
  \end{subfigure}
  \begin{subfigure}[b]{0.24\textwidth}
    \includegraphics[width=\textwidth]{images/subpixel}
  \end{subfigure}\\[\medskipamount]
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  framerate_equalizer {}
  compound {
    channel "source1"
    phase 0 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source2"
    phase 1 period 3
    outputframe  { name "frame" }
  }
  compound {
    channel "source3"
    phase 2 period 3
    outputframe  { name "frame" }
  }
  inputframe { name "frame" }
}
    \end{lstlisting}\vspace{24ex}}
    \caption{\label{fig:dplex}Time-Multiplex}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "destination"
  outputtiles {
    name "queue"
    size [ 64 64 ]
  }
  compound {
    channel "destination"
    inputtiles { name "queue" }
  }
  compound {
    channel "source1"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source2"
    inputtiles { name "queue" }
    outputframe {}
  }
  compound {
    channel "source3"
    inputtiles { name "queue" }
    outputframe {}
  }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
  inputframe { name "frame.source3" }
}
    \end{lstlisting}}
    \caption{\label{fig:tiles}Tiles}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    pixel [ 0 0 3 1 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    pixel [ 1 0 3 1 ]
    outputframe {}
  }
  compound {
    channel "source2"
    pixel [ 2 0 3 1 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:pixel}Pixel}
  \end{subfigure}
  \begin{subfigure}[t]{0.24\textwidth}
    {\tiny\begin{lstlisting}
compound {
  channel "dest"
  compound {
    channel "dest"
    subpixel [ 0 3 ]
    outputframe { type texture }
  }
  compound {
    channel "source1"
    subpixel [ 1 3 ]
    outputframe {}
  }
  compound {
    channel "source2"
    subpixel [ 1 3 ]
    outputframe {}
  }
  inputframe { name "frame.dest" }
  inputframe { name "frame.source1" }
  inputframe { name "frame.source2" }
}
    \end{lstlisting}\vspace{21ex}}
    \caption{\label{fig:subpixel}Subpixel}
  \end{subfigure}
  \caption{New Equalizer task decompositions and their compounds for parallel
    rendering}
  \label{fig:compounds}
\end{figure*}

@@

Time-multiplexing (\fig{fig:dplex}), also called AFR or DPlex, was first
implemented in \cite{BRE:05} for shared memory machines. It is however a better
fit for distributed memory systems, since the separate memory space makes
concurrent rendering of different frames easier to implement. While it increases
the framerate linearly, it does not decrease the latency between user input and
the corresponding output. Consequently, this decomposition mode is mostly useful
for non-interactive movie generation. It is transparent to Equalizer
applications, but does require the configuration latency to be equal or greater
than the number of source channels. Furthermore, to work in a multi-threaded,
multi-GPU configuration, the application needs to support running the rendering
threads asynchronously, as outlined in \sref{sec:threading}. The output frame
rate of the destination channel is smoothened using a frame rate equalizer
(\sref{sec:framerateEq}).

@@

\subsubsection{Frame Rate Equalizer}\label{sec:framerateEq}

@@

\subsubsection{Thread Synchronization Modes}\label{sec:threading}

@@

Head tracking is the minimal feature needed to support immersive
installations. Equalizer does support multiple, independent tracked views
through the observer abstraction introduced in \sref{sec:observer}. Built-in
VRPN support enables the direct, application-transparent configuration of a VRPN
tracker device. Alternatively, an application can provide a $4\times 4$ tracking
matrix defining the transformation from the canvas coordinate system to the
observer. Both CAVE-like tracking with fixed projection surfaces and HMD
tracking modes are implemented.

@@

Applications can switch each view between mono and stereo rendering at runtime,
and can run both monoscopic and stereoscopic views concurrently
(\fig{FIG_teaser} (b)). While this is trivial for active stereo rendering, it
does potentially involve the start and stop of resources and processes for
passive stereo or stereo-dependent task decompositions, as described in
\sref{sec:reconfig}.

@@

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved
screens. These frusta describe a physically-correct display setup for a Virtual
Reality installation. \label{sec:swap} A canvas may have a software or hardware
swap barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a Collage barrier (\sref{sec:barrier}) to synchronize all segments, the
swap buffers call followed by a \textsf{glFlush} to ensure timely execution of
the swap command. The hardware swap barrier is implemented using the
\textsf{WGL\_NV\_swap\_group} or \textsf{GLX\_NV\_swap\_group} extensions.

@@

\label{sec:observer}
An observer looks at one or more views. It is described by the observer position
in the world and its eye separation. Each observer will have its own stereo
mode, focus distance and frame loop (framerate). This allows to have untracked
views and multiple tracked views, e.g., two HMDs, in the same application.

\subsection{Runtime Reconfiguration}\label{sec:reconfig}

@@

\begin{compactdesc}
\item[Connection] A stream-oriented point-to-point communication
  line. The connections
  transmit raw data reliably between two endpoints for unicast connections, and
  between a set of endpoints for multicast connections. For unicast,
  process-local pipes, TCP and Infiniband RDMA are implemented. For multicast,
  a reliable, UDP-based protocol is discussed in \sref{sec:RSP}.
\item[DataI/OStream] Abstracts the input and output of C++ data types from or to
  a set of connections by implementing output stream operators. Uses buffering
  to aggregate data for network transmission. Performs endian swapping during
  input if the endianness differs between the originating and local node.
\item[Node and LocalNode] The abstraction of a process in the cluster. Nodes
  communicate with each other using connections. A LocalNode listens on various
  connections and processes requests for a given process. Received data is
  wrapped in ICommands and dispatched to command handler methods. A Node is a
  proxy for a remote LocalNode. The Equalizer Client object is a LocalNode.
\item[Object] Provides object-oriented, versioned data distribution of C++
  objects between nodes. Objects are registered or mapped on a Local\-Node.
\end{compactdesc}

@@

\subsection{Reliable Stream Protocol}\label{sec:RSP}

@@

\label{sec:Serializable}The Collage Serializable implements one convenient usage
pattern for object data distribution. The \textsf{co::Serializable} data
distribution is based on the concept of dirty bits, allowing inheritance with
data distribution. Dirty bits form a 64-bit mask which marks the parts of the
object to be distributed during the next commit. For serialization, the
application developer implements \textsf{serialize} or \textsf{deserialize},
which are called with the bit mask specifying which data has to be transmitted
or received. During a commit or sync, the current dirty bits are given, whereas
during object mapping all dirty bits are passed to the serialization methods.

@@

\subsection{Barriers, Queues and Object Maps}\label{sec:barrier}

@@

Livre (Large-scale Interactive Volume Rendering Engine) is an out-of-core
parallel 4D volume renderer. It supports sort-first decomposition, both for
scalability as well as for driving large-scale tiled display walls. The
out-of-core loading of data relies on a shared file system, only the high-level
state, e.g., camera position and rendering settings, are shared through Collage
objects.

@@

RTT Deltagen (now Dassault 3D Excite) is a commercial application for
interactive, high quality rendering of CAD data. The RTT Scale module,
delivering multi-GPU and distributed execution, is based on Equalizer and
Collage, and has driven many of the aforementioned features.

RTT Scale uses a master-slave execution mode, were a single running Deltagen
instance can go into "Scale mode" at any time by launching an Equalizer
configuration. Consequently, the whole internal representation needed for
rendering is based on a Collage-based data distribution. The rendering clients
are separate, smaller applications which will map their scenes during
startup. At runtime, any change performed in the main application is committed
as a delta at the beginning of the next frame, following a design pattern
similar to the Collage \textsf{Serializable}
(\sref{sec:Serializable}). Multicast (\sref{sec:RSP}) is used to keep data
distribution times during session launch reasonable for larger cluster sizes
(tens to hundreds of nodes).

RTT Scale is used for a wide variety of use cases. In virtual reality, the
application is used for virtual prototyping and design reviews in front of
high-resolution display walls or in a CAVE and for virtual prototyping of
human-machine interactions in a CAVE or using HMDs. For scalability, sort-first
and tile compounds are used to achieve fast, high-quality rendering, primarily
for interactive raytracing, both based on CPUs and GPUs. For CPU-based
raytracing, often Linux-based rendering clients are used with a Windows-based
application node.

@@

%----------------------------------------------------------------------
\section{Experimental Results}
%----------------------------------------------------------------------
\label{sec:results}

@@

Distribute a video texture (per frame)

@@

%----------------------------------------------------------------------
\section{Discussion and Conclusion}
%----------------------------------------------------------------------
\label{sec:conclusions}

@@

We would also like to thank all supporters and contributors of Equalizer, most
notably RTT, the Blue Brain Project, the University of Siegen, EVL, Dardo, TBD.
